---
title: "BellaBeat's Competitor's Performance"
author: "Anorup Kanti Das"
date: "2025-08-27"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About The Analysis
This analysis explores data from Fitbit, one of Bellabeat’s key competitors in the smart device market. By examining user activity and behavior trends, we aim to uncover insights that can guide Bellabeat’s marketing efforts. Understanding how consumers engage with Fitbit devices will help us identify opportunities to better position Bellabeat and tailor our messaging to appeal to health-conscious users.


#### About The Data.
The data belongs to Fitbit Fitness Tracker data. You can download the datasets [here](https://www.kaggle.com/datasets/arashnic/fitbit). Although the data has some limitations such as not being current, the sample size being small and not gender specific. Despite these limitations, insights remain valuable for understanding general health trends. We will still work with as much data we have to analyze and derive conclusions. After that we will share recommendations to Bellabeat's marketing team to promote their products.

```{r include=FALSE, warning=FALSE, message=FALSE}
# Loading the tidyverse packages
library(tidyverse)
library(lubridate) # For working with dates
library(janitor)   # For cleaning column names

# Importing the main data files we need
daily_activity <- read_csv("dailyActivity_merged.csv")
sleep_day <- read_csv("sleepDay_merged.csv")
hourly_steps <- read_csv("hourlySteps_merged.csv")
```

```{r include=FALSE, warning=FALSE, message=FALSE}
daily_activity <- clean_names(daily_activity)
sleep_day <- clean_names(sleep_day)
hourly_steps <- clean_names(hourly_steps)

# Converting date columns to the correct date format
daily_activity$activity_date <- mdy(daily_activity$activity_date)
sleep_day$sleep_day <- mdy_hms(sleep_day$sleep_day)

# Adding a day of the week column to see trends
daily_activity <- daily_activity %>%
  mutate(day_of_week = wday(activity_date, label = TRUE, abbr = FALSE))

# Mergeing the dataframes to analyze sleep and activity together
merged_data <- left_join(daily_activity, sleep_day, by = c("id" = "id", "activity_date" = "sleep_day"))
```

```{r include=FALSE, warning=FALSE, message=FALSE}
daily_trends <- daily_activity %>%
  group_by(day_of_week) %>%
  summarise(
    avg_steps = mean(total_steps),
    avg_calories = mean(calories)
  )

```

## Analysis & Visualizations
<br>

#### 1. Analyze Daily Activity Trends

Find the average number of steps and calories for each day of the week.
```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = daily_trends, aes(x = day_of_week, y = avg_steps, fill = day_of_week)) +
  geom_col() +
  labs(
    title = "Average Steps Per Day of the Week",
    x = "Day of the Week",
    y = "Average Steps"
  ) +
  theme_minimal()
```
<br>
* **Insight**: This analysis often shows that users are less active on weekends (especially Sundays) and most active during the middle of the week. This is a key finding for Bellabeat's marketing.

<br>
<br>

#### 2 Analyze the Relationship Between Steps and Calories
See if a person's step count is related to the calories they burn.
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = daily_activity, aes(x = total_steps, y = calories)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Steps and Calories Burned",
    x = "Total Steps",
    y = "Calories Burned"
  ) +
  theme_minimal()
```
<br>
* **Insight**: This plot will show a strong positive correlation. This is a good piece of information for Bellabeat, as it confirms that their devices accurately track activity and calorie expenditure.

<br>
<br>

#### 3. Analyze User Sleep Habits
Look at the data to see how much users are sleeping on average.

```{r include=FALSE, warning=FALSE, message=FALSE}
# Find the average minutes asleep
avg_sleep_minutes <- mean(sleep_day$total_minutes_asleep, na.rm = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot a histogram to show the distribution of sleep
ggplot(data = sleep_day, aes(x = total_minutes_asleep / 60)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of User Sleep Duration",
    x = "Hours Asleep",
    y = "Number of Records"
  ) +
  geom_vline(xintercept = 7, color = "red", linetype = "dashed", size = 1) +
  theme_minimal()
```
<br>

* **Insight**: The average user sleeps around 7 hours, which is a good amount. However, the histogram shows that there's a wide range, with many users sleeping much less than the recommended 7-9 hours. This is a perfect opportunity for BellaBeat.

<br>
<br>

## Recommendations For Bellabeat
Based on the insights, here are three simple, high-level recommendations for Bellabeat's marketing team.

#### 1. Create "Weekend Wind-Down" Campaigns

* **Insight**: Users are least active on weekends.

* **Recommendation**: Encourage consistent device usage through weekend-focused marketing. Bellabeat can send push notifications on Fridays or Saturdays with ideas for mindful, low-intensity activities like yoga or a relaxing nature walk, positioning the Leaf as a tool for recovery and well-being.

#### 2. Highlight the Importance of Sleep

* **Insight**: While users track steps, many don't consistently track their sleep, and some get less than the recommended amount.

* **Recommendation**: Launch a campaign called "Sleep & Shine." Focus on promoting the importance of sleep for beauty and wellness. Use the Bellabeat app to send reminders to users to wind down before bed and showcase the app's sleep-tracking features.

#### 3. Use Gamification to Drive Engagement

* **Insight**: The most active users are consistently taking steps and burning calories.

* **Recommendation**: Bellabeat can increase engagement by creating weekly step or active minutes challenges. Reward users with virtual badges or small discounts for hitting their goals. This can motivate users to be more active and to use the product more consistently.